Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-27 paper/paper.tex
fixes throughout

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally back to direct-style with delimited control operators. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the first-order datatypes representing continuations
  into higher-order functions. 
  The refunctionalized AAM is an abstract interpreter written in continuation-passing style
  with two layers of continuation.

@@

Defining a language by building an interpreter for it can be traced to the very
early days of programming languages research \cite{landin1966next, Reynolds:72, 
McCarthy:1960:RFS:367177.367199}.
Nowadays, even undergraduate students in computer science are able to make their own
programming language by building an interpreter.
But building a sound abstract interpreter remained an esoteric and difficult task 
until very recently.

@@

In the domain of abstract semantic artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the 
question of deriving one from the other, are not well studied.
One of the fundamental differences between concrete semantic artifacts and abstract semantic
artifacts is a sound latter one must tell you facts about all possible execution paths, and 
thus nondeterminism is introduced. In addition to ensuring termination, abstract semantic
artifacts are usually equipped with a cache of reachable states, the cache is
updated in a monotonic way and guaranteed to reach the least fixed-point eventually.
In this functional pearl, we make a first step to answering the question of interderiving
abstract semantics artifacts, and provide a constructive answer to relate AAM and ADI
by presenting a series of syntactical transformations on the program from small-step 
abstract abstract machines to big-step abstract definitional interpreters.

@@

\begin{figure}
\begin{center}
 \begin{tikzpicture}
   \matrix (m) [matrix of math nodes, row sep=2.5em, column sep=3.7em]
     {
         &  &  &  &
       \begin{smallmatrix} \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
       \begin{smallmatrix} \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
       \begin{smallmatrix} \text{disentangled}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{refunctionalized}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{monadic} \\ \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix}  \\
       \begin{smallmatrix} \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
         &
       %\begin{smallmatrix} \text{disentangled} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{refunctionalized} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
     };
   {
     [start chain] \chainin (m-2-4);
     \chainin (m-1-5) [double,join={node[sloped,anchor=center,above,labeled] {\text{CPS left-inv.}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{}}}];
     \chainin (m-1-5) [join={node[above,labeled] {\text{}}}];
   }
   %{ [start chain] \chainin (m-1-5);
   % \chainin (m-2-4) [join={node[sloped,anchor=center,above,labeled] {\text{CPS trans.}}}]; }
   %
   % TODO: find out how to make the arrows visually separate
   % {
   %   [start chain] \chainin (m-1-5);
   %   \chainin (m-2-4) [bend left=5, join={node[sloped,anchor=center,below,labeled] {\text{CPS}}}];
   % }
   {
     [start chain] \chainin (m-2-1);
     \chainin (m-2-2) [join={node[above,labeled] {\begin{smallmatrix} \text{linearization} \\ \text{+} \\ \text{fusion} \end{smallmatrix}}     }];
     \chainin (m-2-3) [join={node[above,labeled] {\text{disentang.}}}];
     \chainin (m-2-4) [join={node[above,labeled] {\text{refunc.}}}];
     %\chainin (m-2-5) [join={node[above,labeled] {\text{monadify}}}];
   }
   {
     [start chain, color=gray]
     \chainin (m-3-1);
     { [start branch=A] \chainin (m-2-1)
       [join={node[left, labeled] {\alpha}}];
     }
   }
     %\chainin (m-2-2);
     %\chainin (m-2-3);
     %\chainin (m-2-4);
   {
     [start chain, color=gray]
     \chainin (m-3-5);
     { [start branch=B] \chainin (m-2-5)
       [join={node[right, labeled] {\alpha}}];
     }
     {
       [start branch=D, color=gray] \chainin (m-3-1)
       [join={node[right, labeled] {\text{}}}];
     }
   }
   {
     [start chain, color=gray] \chainin (m-3-1);
     {
       [start branch=C] \chainin (m-3-5)
       [join={node[above, labeled] {\text{Known as Danvy's correspondences for concrete semantic artifacts }}}];
     }
   }
 \end{tikzpicture}
 \caption{Transformations from AAM to ADI} \label{fig:trans}
\end{center}
\end{figure}

@@

We begin by reviewing some background necessary for this work in Section~\ref{background}, 
as well as introducing some of the basic code structures used throughout the paper. 
We then address the main contribution of this paper, which is the filling of the gap between
small-step AAM and big-step ADI by applying a series of well-known systematic transformations
drawn from concrete semantic artifacts.
Figure \ref{fig:trans} shows the transformations.

Those transformations are summarized here, with their associated section:

@@

  \item Section~\ref{disen} discusses \textbf{disentanglement}. In this part, we identify 
    different first-order datatypes which represent different layers of continuation, and
    disassemble their handlers to be separate higher-order functions.

@@

There are two main reasons we use a real-world language:
1) The code does not diminish the accuracy of the material than formal and mathematical
notations, which are heavily used in other static analysis or semantics papers.
The Scala code in this paper can be easily back-translated into formal notations.
2) As a functional pearl, the code in this paper is executable with only few changes,
which make it particularly fit for presenting syntactical transformations on abstract
interpreters.
\fi

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines
and big-step abstract definitional interpreters by applying a series of syntactical
transformations. Among these transformations, linearization turns a worklist into another
layer of continuations; refunctionalization converts the first-order data types representing
continuations to higher-order functions; and finally a left-inverse CPS transformation turns
the abstract interpreter into a sequence of expressions with delimited control operators,
which looks almost no different to a concrete interpreter.

@@

%% TODO: Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

>>>>

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally transformation from CPS back to direct-style 
  with delimited control operators. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the first-order datatypes representing continuations
  into higher-order functions. 
  The refunctionalized AAM is an abstract interpreter written in continuation-passing style
  with two layers of continuation.

@@

Defining a language by building an interpreter for it can be traced to the very
early days of programming languages research \cite{McCarthy:1960:RFS:367177.367199,landin1966next, Reynolds:72}.
Nowadays, even undergraduate students in computer science learn how to make their own
programming languages by building an interpreter.
But building a sound abstract interpreter remained an esoteric and difficult task 
until very recently.

@@

In the domain of abstract semantic artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the 
question of deriving one from the other, are not well studied.
One of the fundamental differences between concrete semantic artifacts and abstract semantic
artifacts is a sound latter one must tell you facts about all possible execution paths, and 
thus nondeterminism is introduced. In addition to ensuring termination, abstract semantic
artifacts are usually equipped with a cache of reachable states, the cache is
updated in a monotonic way and guaranteed to reach the least fixed-point eventually.
In this functional pearl, we make a first step to answering the question of interderiving
abstract semantics artifacts, and provide a constructive answer to relate AAM and ADI 
by presenting a series of syntactic transformations on the program from small-step 
abstract abstract machines to big-step abstract definitional interpreters.

@@

\begin{figure}
\begin{center}
 \begin{tikzpicture}
   \matrix (m) [matrix of math nodes, row sep=2.5em, column sep=3.7em]
     {
         &  &  &  &
       \begin{smallmatrix} \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
       \begin{smallmatrix} \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
       \begin{smallmatrix} \text{disentangled}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{refunctionalized}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{monadic} \\ \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix}  \\
       \begin{smallmatrix} \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
         &
       %\begin{smallmatrix} \text{disentangled} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{refunctionalized} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
     };
   {
     [start chain] \chainin (m-2-4);
     \chainin (m-1-5) [double,join={node[sloped,anchor=center,above,labeled] {\text{CPS left-inv.}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{}}}];
     \chainin (m-1-5) [join={node[above,labeled] {\text{}}}];
   }
   %{ [start chain] \chainin (m-1-5);
   % \chainin (m-2-4) [join={node[sloped,anchor=center,above,labeled] {\text{CPS trans.}}}]; }
   %
   % TODO: find out how to make the arrows visually separate
   % {
   %   [start chain] \chainin (m-1-5);
   %   \chainin (m-2-4) [bend left=5, join={node[sloped,anchor=center,below,labeled] {\text{CPS}}}];
   % }
   {
     [start chain] \chainin (m-2-1);
     \chainin (m-2-2) [join={node[above,labeled] {\begin{smallmatrix} \text{linearization} \\ \text{+} \\ \text{lightw. fusion} \end{smallmatrix}}     }];
     \chainin (m-2-3) [join={node[above,labeled] {\text{disentang.}}}];
     \chainin (m-2-4) [join={node[above,labeled] {\text{refunc.}}}];
     %\chainin (m-2-5) [join={node[above,labeled] {\text{monadify}}}];
   }
   {
     [start chain, color=gray]
     \chainin (m-3-1);
     { [start branch=A] \chainin (m-2-1)
       [join={node[left, labeled] {\alpha}}];
     }
   }
     %\chainin (m-2-2);
     %\chainin (m-2-3);
     %\chainin (m-2-4);
   {
     [start chain, color=gray]
     \chainin (m-3-5);
     { [start branch=B] \chainin (m-2-5)
       [join={node[right, labeled] {\alpha}}];
     }
     {
       [start branch=D, color=gray] \chainin (m-3-1)
       [join={node[right, labeled] {\text{}}}];
     }
   }
   {
     [start chain, color=gray] \chainin (m-3-1);
     {
       [start branch=C] \chainin (m-3-5)
       [join={node[above, labeled] {\text{Known as Danvy's correspondences for concrete semantic artifacts }}}];
     }
   }
 \end{tikzpicture}
 \caption{Transformations from AAM to ADI} \label{fig:trans}
\end{center}
\end{figure}

@@

We begin by reviewing some background necessary for this work in Section~\ref{background}, 
as well as introducing some of the basic code structures used throughout the paper. 
We then address the main contribution of this paper: bridging the gap between
small-step AAM and big-step ADI by applying a series of well-known systematic transformations
drawn from concrete semantic artifacts.
Those transformations are shown in Figure~\ref{fig:trans} and summarized here, with their 
respective associated sections:

@@

  \item Section~\ref{disen} discusses \textbf{disentanglement}. In this part, we identify 
    different first-order datatypes which represent different layers of continuations, and
    disassemble their handlers into separate higher-order functions.

@@

There are two main reasons we use a real-world language:
1) The code does not diminish the accuracy of the material than formal and mathematical
notations, which are heavily used in other static analysis or semantics papers.
The Scala code in this paper can be easily back-translated into formal notations.
2) As a functional pearl, the code in this paper is executable with only few changes,
which make it particularly fit for presenting syntactic transformations on abstract
interpreters.
\fi

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines
and big-step abstract definitional interpreters by applying a series of syntactic 
transformations. Among these transformations, linearization turns a worklist into another
layer of continuations; refunctionalization converts the first-order data types representing
continuations to higher-order functions; and finally a left-inverse CPS transformation turns
the abstract interpreter into a sequence of expressions with delimited control operators,
which looks almost no different to a concrete interpreter.

@@

%% TODO: Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  The authors are grateful to the anonymous reviewers for their 
  helpful comments, and they are further indebted to Olivier Danvy 
  for numerous invaluable comments on the draft.

  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the authors and do not necessarily reflect the views of the
  National Science Foundation. \note{TODO: grants}
\end{acks}
